---
title: "Summary of bird response to grazing"
output: word_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      error = FALSE)
```



```{r}

library(tidyverse)
library(lubridate)
library(ggrepel)
library(flextable)
library(here)
library(birdnames)

source(here("code/bird_functions.R"))

```

```{r}

cgrc_birds <- read.csv(here("data/CGRC_AS.csv")) %>% 
  clean_area_search() %>% 
  bird_taxa_filter(keep_taxa = c("Passeriformes", "Apodiformes", "Cathartiformes", "Galliformes", "Piciformes"))



birds_area_year <- cgrc_birds %>% 
  birds_per_area_year() %>% 
  group_by(point, year) %>% 
  summarise(total.birds = sum(mean.birds)) %>% 
  ungroup() %>% 
  full_join(read.csv(here("data/treatments.csv")))  %>% 
  mutate(treatment = ifelse(point == "CGRC-02", "Grazed", "No treatment"))


species_area_year <- cgrc_birds %>% 
  species_per_area_year() %>% 
  full_join(read.csv(here("data/treatments.csv"))) %>%
  mutate(treatment = ifelse(point == "CGRC-02", "Grazed", "No treatment"))



```

# Did we see any change in bird abundance or species diversity after the first year of prescribed grazing?  

Below is a brief summary of the bird data from the first post-treatment year of our prescribed disturbance project.

We surveyed birds in all plots in both years. Thus, in summer 2023 we have data from 9 untreated plots to compare to our one treated plot. I calculated the total number of individual birds of all species combined, and the total number of species detected at each plot each year. I then calculated the average number of birds or species per plot and year. In all figures below, for untreated plots, the light lines show the change for each untreated plot, the heavy red line is the average of all untreated plots, and the heavy green line is the change in the treated plot.  

Overall, there isn't a clear overall treatment effect evident for either number of individual birds or number of bird species. In both cases, the slopes of the heavy lines are similar between treated and untreated plots. There were more individual birds and more species detected in untreated plots, but untreated represents a much larger area and these values are not corrected for area.


```{r}
birds_area_year %>% 
  ggplot() +
  geom_line(stat="smooth", aes(x = year, y = total.birds, group = point, color = treatment), alpha = 0.5, method = "lm") +
  geom_point(aes(x = year, y = total.birds, color = treatment), size = 2) +
  stat_smooth(aes(x = year, y = total.birds, color = treatment), size = 2, method = "lm", se = FALSE) +
  expand_limits(y = 0) +
  geom_text_repel(data = filter(birds_area_year, 
                                year == 2022), aes(label = point, x = 2022, y = total.birds), nudge_x = -.2, size = 2.5) +
  scale_x_continuous(breaks = seq(min(birds_area_year$year), max(birds_area_year$year))) +
  scale_color_brewer(palette = "Dark2") +
  theme_bw() +
  labs(color = NULL,
       x = "",
       y = "Mean # birds detected per survey",
       title = "Number of individual birds detected in each plot") +
  theme(legend.position=c(.5,.2))



```

```{r}

species_area_year %>% 
  ggplot() +
  geom_line(stat="smooth", aes(x = year, y = mean.species, group = point, color = treatment), alpha = 0.5, method = "lm") +
  geom_point(aes(x = year, y = mean.species, color = treatment), size = 2) +
  stat_smooth(aes(x = year, y = mean.species, color = treatment), size = 2, method = "lm", se = FALSE) +
  expand_limits(y = 0) +
  geom_text_repel(data = filter(species_area_year, 
                                year == 2022), aes(label = point, x = 2022, y = mean.species), nudge_x = -.2, size = 2.5) +
  scale_x_continuous(breaks = seq(min(species_area_year$year), max(species_area_year$year))) +
  scale_color_brewer(palette = "Dark2") +
  theme_bw() +
  labs(color = NULL,
       x = "",
       y = "Mean # species detected per survey",
       title = "Number of bird species detected at each plot") +
  theme(legend.position=c(.5, .15))


```


```{r}
spp_table <- cgrc_birds  %>% 
  mutate(year = year(date)) %>% 
  full_join(read.csv(here("data/treatments.csv"))) %>% 
  mutate(treatment = ifelse(point == "CGRC-02", "Grazed", "No treatment")) 

spp_table_wide <- spp_table %>% 
  distinct(treatment, common.name, before.after) %>% 
  mutate(treatment.time = paste(treatment, before.after, sep = ", ")) %>% 
  pivot_longer(cols = treatment.time) %>% 
  mutate(detected = "Detected") %>% 
  pivot_wider(id_cols = c("common.name"), names_from = value, values_from = detected) %>% 
  arrange(common.name)

spp_table_long <- spp_table_wide %>% 
  pivot_longer(-common.name, names_to = "treatment.time", values_to = "detected") %>% 
  separate(treatment.time, c("treatment", "before.after"), remove = FALSE, sep = ", ") %>% 
  mutate(detected = replace_na(detected, "Not detected"),
         treatment.time = gsub(", ", ",\n", treatment.time),
         treatment.time = factor(treatment.time, levels = c("No treatment,\nbefore", "No treatment,\nafter", "Grazed,\nbefore", "Grazed,\nafter")),
         before.after = factor(before.after, levels = c("before", "after")))

```

When we look at presence vs absence of each species between grazed and ungrazed plots before and after treatment, we perhaps gain some additional insight, although the small sample size caveat must not be forgotten. Nevertheless, we did see some species-specific responses to grazing that might have been expected, notably losing Bewick's Wren and Bushtit, which favor more dense shrubby habitats, and gaining White-crowned Sparrow, which favors more open grassy habitats with sparser shrubs. 



```{r fig.width=8, fig.height=10.5}
spp_table_long %>% 
  ggplot() +
  geom_tile(aes(y = common.name, x = before.after, fill = detected), color = "gray25") +
  theme_bw() +
  labs(x = "",
       y = "",
       fill = NULL,
       title = "Bird species detections in grazed and untreated plots\nbefore and after treatment") +
  scale_fill_brewer(type = "qual", palette = "Set1", direction = -1) +
    theme(legend.position="bottom") +
  facet_wrap(~treatment, scales = "free_x", ncol = 2) +
    scale_y_discrete(limits=rev)

```


```{r eval = FALSE}

spp_table %>% 
  flextable() %>% 
  autofit()

```

