---
title: "Prescribed disturbance plant response"
output: word_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE, 
                      fig.width=7)
```


```{r}


library(tidyverse)
library(stringi)
library(here)
library(ggrepel)
library(ggbeeswarm)
library(flextable)


source(here("code/general_functions.R"))
source(here("code/veg_functions.R"))
options(scipen = 999)

zwidth = 0.05

```



```{r}


percent_cover <- readRDS(here("data/derived/percent_cover"))

richness <- readRDS(here("data/derived/richness"))

points_years_treatments <- readRDS(here("data/derived/points_years_treatments"))  %>% 
  rename(year = treatment.effect.year) %>% 
  select(point, year, treatment)%>% 
  mutate(treat.lump = ifelse(treatment %in% c("none", "before treatment"), "untreated", treatment))

points_years_monitoring <- percent_cover %>% 
  distinct(point, year)

drop_points <- c("CGRC-07", "CGRC-08")

# treated point and label position jitter
pos <- position_jitter(height = 0.1, seed = 2)

```



```{r include = FALSE}

treatment_hx <- points_years_treatments %>% 
  group_by(point, treatment) %>% 
  mutate(treat.year = row_number()) %>% 
  ungroup() %>% 
  mutate(zyears = ifelse(treat.year == 1, "year", "years"),
         treat.hx.year = ifelse(!treatment %in% c("none", "before treatment"), paste(treatment, ", ", treat.year, " ", zyears, sep = ""), treatment),
         treat.hx.year = factor(treat.hx.year, levels = c("none", "before treatment", "grazed, 1 year", "grazed, 2 years", "hand clear, 1 year"))
         ) 
  

```
  
  We are now in the third year of our prescribed disturbance study at the Cypress Grove Preserve. In the first year we collected baseline vegetation data at 7 of 10 study plots and did not do any disturbance treatments. In the fall of 2022, we grazed plot CGRC-02, and in fall of 2023 we grazed plot CGRC-02 again and also grazed plot CGRC-04. In early spring, before vegetation monitoring, we also hand cleared shrubby vegetation from plot CGRC-05. 

```{r}

ggplot() +
  geom_tile(data = points_years_treatments, aes(y = reorder(point, desc(point)), x = year, fill = treatment), color = "black") +
  scale_fill_manual(values = c("#4DAF4A", "#E41A1C", "#377EB8", "#999999")) +
  #scale_fill_brewer(palette = "Dark2") +
  labs(title = "Treatment history at each point",
       x = "",
       y = "",
       fill = "Treatment",
       caption = "Areas without color indicate no vegetation monitoring for that plot/year.") +
  theme_minimal()

```

We have also done standard vegetation monitoring in each of these study plots, except that we only monitored plots CGRC-07 and CGRC-08 in 2022, and we did not monitor CGRC-04 in 2023. We then calculated percent cover and species richness to evaluate the effectiveness of our stewardship efforts. There is substantial interannual variability in cover and richness in grasslands, so evaluating the trajectory of each point through time may not be informative. Instead, we can estimate the overall variability in the untreated points and years, then compare treated points against that variability. If treated points are consistently on one side or the other of the mean untreated value, then we have some evidence that our stewardship is having a certain effect. If treated points are consistently above or below the normal observed range of untreated values, then we may conclude that we have good evidence that our stewardship is having a given effect.

In our study area we have detected no Native annual grasses and no Invasive perennial forbs so those panels are blank in the figures below. The monitoring transects in plot CGRC-05 did not intersect the area that was hand cleared so that plot is not shown below.  
  
Percent cover 

```{r}
per_cov_treat <- percent_cover %>% 
  filter(!point %in% drop_points, cover.type == "abs.cover") %>%
  full_join(points_years_treatments) 

mean_untreated_per_cover <- per_cov_treat %>% 
  filter(treat.lump == "untreated") %>% 
  group_by(FunGrp, nat.nnat.inv) %>% 
  summarise(mean.cov = mean(cover),
            sd.cov = sd(cover)) %>% 
  ungroup()

#  filter(!(FunGrp == "Annual Grass" & nat.nnat.inv == "Native"), !(nat.nnat.inv == "Invasive" & FunGrp == "Perennial Forb"))





```


Forbs  
Grazing appears to have increased native annual forb percent cover in both treated plots and may also have increased native perennial forb cover in both plots.  

  
```{r fig.height = 4}

fungrp_cover_plotter <- function(zfungrp) {
fungrp_df <- filter(per_cov_treat, str_detect(FunGrp, zfungrp))
mean_untreated <- filter(mean_untreated_per_cover, str_detect(FunGrp, zfungrp))
  
cov_plot <- ggplot() +
  geom_rect(data = mean_untreated, aes(ymax = Inf, ymin = -Inf, xmin = mean.cov - sd.cov, xmax = mean.cov + sd.cov), fill = "gray", alpha = 0.25) +
  geom_vline(data = mean_untreated, aes(xintercept = mean.cov)) +
  geom_point(data = filter(fungrp_df, treat.lump != "untreated"), aes(y = reorder(treat.lump, desc(treat.lump)), x = cover, color = treatment, shape = point), size = 3, position = pos) +
  geom_beeswarm(data = filter(fungrp_df, treatment == "before treatment"), aes(y = reorder(treat.lump, desc(treat.lump)), x = cover, shape = point, color = treatment), size = 2, dodge.width = 0.75) +
  geom_beeswarm(data = filter(fungrp_df, treatment == "none"), aes(y = reorder(treat.lump, desc(treat.lump)), x = cover, color = treatment), size = 2, dodge.width = 0.75) +
  geom_text_repel(data = filter(fungrp_df, treat.lump != "untreated"),
                  aes(y = reorder(treat.lump, desc(treat.lump)), x = cover, 
                 color = treatment, label = year),
                  size = 4,
                 #min.segment.length = Inf, 
                 position = pos,
                  box.padding   = .5,
                  segment.size = 0.1,
                  segment.color = "grey50")  +
  geom_beeswarm(data = filter(fungrp_df, treatment == "none"), aes(y = reorder(treat.lump, desc(treat.lump)), x = cover), color = "gray30", size = 2, dodge.width = 0.75) +
  theme_bw() +
  facet_grid(FunGrp~nat.nnat.inv, scales = "free_x") +
  labs(x = "% cover",
       y = "",
       shape = "Treated points",
       caption = "Untreated includes points not assigned to treatment and assigned points in years before treatment.\nBlack line and gray area show the mean  ± 1 standard deviation of untreated plots/years.",
       title = "Percent forb cover in treated plots") + 
  scale_color_manual(guide = "none", values = c("#4DAF4A", "#E41A1C", "#377EB8", "#999999")) +
  theme(plot.caption.position = "plot")

ggsave(here(paste("figures/", zfungrp, "_cover_treatments", year(Sys.Date()), ".png", sep = "")), width = 8)

return(cov_plot)

}


```

  
```{r fig.height = 4}
fungrp_cover_plotter("Forb")

```



Grasses  
Grazing appears to have led to strong increases in non-native annual grasses in both treated plots, which is what we expect in the first few years of disturbance. Grazing may also have been associated with increased percent cover for native perennial grasses as well.  


```{r fig.height = 4}

fungrp_cover_plotter("Grass")
```

Shrubs  
Grazing appears to be reducing the cover of shrubby plants in both treated plots.  
  
```{r fig.height = 2.5}


fungrp_cover_plotter("Shrub")


```

Species richness  
Grasses and forbs  
Grazing may be associated with increased number of invasive annual grass species.  

```{r fig.height = 6}
richness_treat <- richness %>% 
  filter(!point %in% drop_points) %>%  
  filter(str_detect(FunGrp, "Grass|Forb")) %>% # grass and forb are the only functional groups with meaningful spp richness
  filter(!(FunGrp == "Annual Grass" & nat.nnat.inv == "Native"), !(nat.nnat.inv == "Invasive" & FunGrp == "Perennial Forb")) %>%  # but there aren't really any native annual grasses or invasive perennial forbs
  left_join(points_years_treatments)
  
  
  mean_untreat_richness <- richness_treat %>% 
    filter(treat.lump == "untreated") %>% 
    group_by(FunGrp, nat.nnat.inv) %>%
    summarise(mean.richness = mean(richness),
              sd.richness = sd(richness)) %>%
    ungroup()

```



```{r}

ggplot() +
  geom_rect(data = mean_untreat_richness, aes(ymax = Inf, ymin = -Inf, xmin = mean.richness - sd.richness, xmax = mean.richness + sd.richness), fill = "gray", alpha = 0.25) +
  geom_vline(data = mean_untreat_richness, aes(xintercept = mean.richness)) +
  geom_point(data = filter(richness_treat, treat.lump != "untreated"), aes(y = reorder(treat.lump, desc(treat.lump)), x = richness, color = treatment, shape = point), size = 3, position = pos) +
  geom_beeswarm(data = filter(richness_treat, treatment == "before treatment"), aes(y = reorder(treat.lump, desc(treat.lump)), x = richness, shape = point, color = treatment), size = 2, dodge.width = 0.75) +
  geom_beeswarm(data = filter(richness_treat, treatment == "none"), aes(y = reorder(treat.lump, desc(treat.lump)), x = richness, color = treatment), size = 2, dodge.width = 0.75) +
  geom_text_repel(data = filter(richness_treat, treat.lump != "untreated"),
                  aes(y = reorder(treat.lump, desc(treat.lump)), x = richness, 
                 color = treatment, label = year),
                  size = 4,
                 #min.segment.length = Inf, 
                 position = pos,
                  box.padding   = .5,
                  segment.size = 0.1,
                  segment.color = "grey50")  +
  theme_bw() +
  facet_grid(FunGrp~nat.nnat.inv, scales = "free_x") +
  labs(title = "Grass and forb species richness in treated plots",
       x = "# of species",
       y = "",
       shape = "Treated points",
       caption = "Untreated includes points not assigned to treatment and assigned points in years before treatment.\nBlack line and gray area show the mean  ± 1 standard deviation of untreated plots/years")   +
  scale_color_manual(guide = "none", values = c("#4DAF4A", "#E41A1C", "#377EB8", "#999999")) +
  theme(plot.caption.position = "plot")


ggsave(here(paste("figures/grass_forb_richness_treatments", year(Sys.Date()), ".png", sep = "")), width = 8)
```



```{r}
spp_list <- readRDS(here("data/derived/long_lpi_assigned")) %>% 
  distinct(Common.Name, Scientific.Name, nat.nnat.inv, FunGrp, point) %>% 
  filter(Common.Name != "") %>% 
  mutate(in.point = "Y",
         FunGrp = replace_na(FunGrp, "Other taxa")) %>%
  pivot_wider(id_cols = c(FunGrp, Common.Name, Scientific.Name, nat.nnat.inv), names_from = point, values_from = in.point) %>% 
  mutate(FunGrp = factor(FunGrp, levels = c("Annual Forb", "Annual Grass", "Perennial Forb", "Perennial Grass", "SedgesRushes", "Legumes", "Shrub", "Tree", "Other taxa"))) %>% 
  arrange(FunGrp, nat.nnat.inv, Scientific.Name) %>% 
  group_by(FunGrp) %>% 
  mutate(spp.num = row_number()) %>% 
  ungroup()

fun_group_nums <- spp_list %>% 
  count(FunGrp) %>% 
  mutate(cum.nums = cumsum(n))
  
spp_list  %>% 
  arrange(FunGrp, nat.nnat.inv, Scientific.Name) %>% 
  mutate(FunGrp = ifelse(spp.num == 1, as.character(FunGrp), "")) %>% 
  select(-spp.num) %>% 
  flextable() %>% 
  set_header_labels(FunGrp = "Functional\ngroup",
                    Common.Name = "Common name",
                    Scientific.Name = "Scientific name",
                    nat.nnat.inv = "Native/Invasive\nstatus") %>% 
  hline(i = fun_group_nums$cum.nums) %>% 
  autofit() %>% 
  fit_to_width(max_width = 8)
  


```

