---
title: "Prescribed disturbance plant response"
output: word_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE, 
                      fig.width=7)
```


```{r}


library(tidyverse)
library(here)
library(ggrepel)
library(flextable)


source(here("code/general_functions.R"))
source(here("code/veg_functions.R"))
options(scipen = 999)

zwidth = 0.05

```



```{r}
treatments <- read.csv(here("data/treatments.csv"))%>% 
  filter(year == 2023) %>% 
  select(-year) %>% 
  mutate(treatment = ifelse(treatment == "control", "none", treatment))


lpi <- read.csv(here("data/CGRC_LPI.csv")) %>% 
  clean_date_year() %>% 
  filter(!is.na(Direction))

long_lpi <- lengthen_lpi(lpi)

long_lpi_assigned <- assign_functional_group_status(long_lpi)

multi_year_points <- lpi %>% 
  split_point_id() %>% 
  distinct(year, point) %>% 
  count(point) %>% 
  filter(n > 1) %>% 
  select(point)



# percent cover ----
veg_cover <- long_lpi_assigned %>% 
  get_fungrp_native_cover() %>% 
  split_point_id()


plot_veg_mean_cov <- veg_cover %>% 
  full_join(expand.grid(year = distinct(veg_cover, year)$year,
                        point = distinct(veg_cover, point)$point,
                        line = distinct(veg_cover, line)$line,
                        FunGrp = distinct(veg_cover, FunGrp)$FunGrp,
                        nat.nnat.inv = distinct(veg_cover, nat.nnat.inv)$nat.nnat.inv)) %>% 
  mutate(abs.cover = ifelse(is.na(abs.cover), 0, abs.cover),
         rel.cover = ifelse(is.na(rel.cover), 0, rel.cover)) %>% 
  pivot_longer(cols = contains("cover"), names_to = "cover.type", values_to = "cover") %>% 
  group_by(year, point, FunGrp, nat.nnat.inv, cover.type) %>% 
  summarise(n.line = n(),
            mean.cov = mean(cover),
            sd.cover = sd(cover)) %>% 
  ungroup() %>% 
  mutate(se.cover = sd.cover/sqrt(n.line),
         lwr.se = mean.cov - se.cover,
         upr.se = mean.cov + se.cover) %>% 
  left_join(treatments)  %>% 
  right_join(multi_year_points)%>% 
  group_by(year, FunGrp, nat.nnat.inv, cover.type, treatment) %>% 
  mutate(grand.mean.cov = mean(mean.cov),
            sd.grand.mean.cov = sd(mean.cov)) %>% 
  ungroup()

```
  
# Did we see any vegetation change after the first year of prescribed grazing?  

From the plant data collected by Point Blue, I calculated the percent cover and species richness of various functional groups of plants. We have data from both years for 4 plots:  

* CGRC-01 is our control plot  
* CGRC-02 was grazed in the late summer of 2022 (after the 2022 veg measurements)  
* CGRC-03 and CGRC-05 are scheduled for treatment in late summer/early fall 2023  
 

Thus, in summer 2023 we have data for three untreated plots to compare to our one treated plot. In all figures below, for untreated plots, the solid lines show the average change across all untreated plots and in the treated plot, and the dashed lines show the change for each untreated plot to give a sense of the variability. 

Overall, you can see that the data are quite noisy, and in particular there was substantial variability in the values calculated between the untreated plots. Thus there isn't a clear overall treatment effect evident in these data. There is some suggestion that both non-native grass and native forb % cover increased meaningfully in the grazed plots. This latter potential result is interesting because there was a decrease in native forb richness in the grazed plot from 3 to only a single species:   

We may expect detectable treatment effects following multiple years of grazing and especially if we can time spring grazing to limit new invasive seeds from being produced. Anecdotally, the sheep apparently had limited capacity/opportunity to browse or knock down most of the larger Coyote brush in the treated plot. Our objectives around woody plant cover will likely need additional disturbance methods.     


```{r}

plot_veg_mean_cov %>% 
  filter(cover.type == "abs.cover", FunGrp == "Shrub", nat.nnat.inv == "Native") %>% 
  ggplot() +
  stat_smooth(aes(x = year, y = mean.cov, color = treatment, group = point), method = "lm", se = FALSE, linetype = 2) +
  geom_point(aes(x = year, y = grand.mean.cov, color = treatment), position=position_dodge(width = zwidth), size = 2) +
  geom_errorbar(aes(x = year, ymin = grand.mean.cov - sd.grand.mean.cov, ymax = grand.mean.cov + sd.grand.mean.cov, color = treatment), position=position_dodge(width = zwidth), width = 0.1) +
  stat_smooth(aes(x = year, y = grand.mean.cov, color = treatment), method = "lm", position=position_dodge(width = zwidth), se = FALSE) +
  scale_x_continuous(labels = c(2022, 2023), breaks = c(2022, 2023), limits = c(2021.9, 2023.15)) +
  geom_text_repel(data = filter(plot_veg_mean_cov, 
                          year == 2023, 
                          treatment == "none", 
                          cover.type == "abs.cover", 
                          FunGrp == "Shrub", 
                          nat.nnat.inv == "Native"), aes(label = point, x = 2023, y = mean.cov), nudge_x = .2, size = 2.5) +
  labs(x = "Year",
       title = "Mean woody plant % cover",
       color = "Treatment",
       y = "% cover") +
  theme_bw()
```




```{r}
plot_fungrp_spec_cover_trt_resp <- function(zFunGrp) {
zdat <- plot_veg_mean_cov %>% 
  filter(cover.type == "abs.cover", FunGrp == zFunGrp)


  ggplot(zdat) +
  stat_smooth(data = filter(zdat, treatment == "none"), aes(x = year, y = mean.cov, color = treatment, group = point), method = "lm", se = FALSE, linetype = 2) +
  geom_point(aes(x = year, y = grand.mean.cov, color = treatment), position=position_dodge(width = zwidth), size = 2) +
  geom_errorbar(aes(x = year, ymin = grand.mean.cov - sd.grand.mean.cov, ymax = grand.mean.cov + sd.grand.mean.cov, color = treatment), position=position_dodge(width = zwidth), width = 0.1) +
  stat_smooth(aes(x = year, y = grand.mean.cov, color = treatment), method = "lm", position=position_dodge(width = zwidth), se = FALSE) +
  scale_x_continuous(labels = c(2022, 2023), breaks = c(2022, 2023), limits = c(2021.9, 2023.5)) +
  geom_text_repel(data = filter(zdat, 
                          year == 2023, 
                          treatment == "none"), aes(label = point, x = 2023, y = mean.cov), nudge_x = .2, size = 2.5) +
  labs(x = "Year",
       y = "% cover",
       color = "Treatment",
       title = paste("Mean", zFunGrp, "% cover")) +
  theme_bw() +
  facet_wrap(~nat.nnat.inv, scales = "free_y", ncol = 2) + 
    expand_limits(y = 0)
}
```


```{r}
plot_fungrp_spec_cover_trt_resp("Annual Grass")
```


```{r}
plot_fungrp_spec_cover_trt_resp("Annual Forb")
```


```{r}
plot_fungrp_spec_cover_trt_resp("Perennial Grass")
```


```{r}
plot_fungrp_spec_cover_trt_resp("Perennial Forb")
```



```{r}


richness <- get_fungrp_native_richness(lpi) %>% 
  left_join(treatments) %>%
  right_join(multi_year_points) %>% 
  group_by(year, FunGrp, nat.nnat.inv, treatment) %>% 
  mutate(mean.richness = mean(richness),
            sd.richness = sd(richness)) %>% 
  ungroup()
  

plot_fungrp_spec_rich_resp <- function(zFunGrp) {
zdat <- richness %>% 
  filter(FunGrp == zFunGrp)
  
  ggplot(zdat) +
  stat_smooth(data = filter(zdat, treatment == "none"), aes(x = year, y = richness, color = treatment, group = point), method = "lm", se = FALSE, linetype = 2) +
  geom_point(aes(x = year, y = mean.richness, color = treatment), position=position_dodge(width = zwidth), size = 2) +
  geom_errorbar(aes(x = year, ymin = mean.richness - sd.richness, ymax = mean.richness + sd.richness, color = treatment), position=position_dodge(width = zwidth), width = 0.1) +
  stat_smooth(aes(x = year, y = mean.richness, color = treatment), method = "lm", position=position_dodge(width = zwidth), se = FALSE) +
  scale_x_continuous(labels = c(2022, 2023), breaks = c(2022, 2023), limits = c(2021.9, 2023.5)) +
  geom_text_repel(data = filter(zdat, 
                          year == 2023, 
                          treatment == "none"), aes(label = point, x = 2023, y = richness), nudge_x = .2, size = 2.5) + 
  labs(x = "Year",
       y = "# species",
       color = "Treatment",
       title = paste("Mean", zFunGrp, "species richness")) +
  theme_bw() +
  facet_wrap(~nat.nnat.inv, scales = "free_y", ncol = 2) + 
    expand_limits(y = 0)
}
```


```{r}
plot_fungrp_spec_rich_resp("Annual Grass")
```

```{r}
plot_fungrp_spec_rich_resp("Annual Forb")
```


```{r}
plot_fungrp_spec_rich_resp("Perennial Grass")
```


```{r}
plot_fungrp_spec_rich_resp("Perennial Forb")
```

```{r}
spp <- read.csv(here("data/CGRC_plant_species_list.csv")) %>% 
  mutate(year = year(as.Date(Event.Date))) %>% 
  split_point_id() %>% 
  distinct(point, year, USDA.Code) %>%
  left_join(read.csv(here("data/helper_files/CAPlantsv2.csv"))%>% dplyr::select('USDA.Code' = Symbol, Scientific.Name, Common.Name, Native.Status, Invasive, FunGrp1, FunGrp2, FunGrp, Growth.Habit)) %>% 
    mutate(nat.nnat.inv = case_when(grepl("L48 \\(N\\)", Native.Status) ~ "Native",
                                    grepl("L48 \\(I\\)", Native.Status) & !grepl("Cal-IPC", Invasive)~ "Non-native",
                                    grepl("L48 \\(I\\)", Native.Status) & grepl("Cal-IPC", Invasive) ~ "Invasive"),
           FunGrp = case_when(FunGrp == "ShrubsTrees" & grepl("hrub", Growth.Habit) ~ "Shrub",
                              FunGrp == "ShrubsTrees" & !grepl("hrub", Growth.Habit) ~ "Tree",
                              TRUE ~ as.character(FunGrp)))
  
spp %>% 
  right_join(multi_year_points) %>% 
  mutate(observed = "Y",
         point.year = paste(point, year, sep = ", "),
         species = paste(Common.Name, " (", Scientific.Name, ")", sep = "")) %>% 
  pivot_wider(id_cols = c(species, FunGrp, nat.nnat.inv), names_from = point.year, values_from = observed) %>% 
  arrange(FunGrp, nat.nnat.inv) %>% 
  flextable() %>% 
  add_header_row(values = c("", "CGRC-01", "CGRC-02", "CGRC-03", "CGRC-05"), colwidths = c(3, 2, 2, 2, 2)) %>% 
  set_header_labels(species = "Species",
                    FunGrp = "Functional\ngroup",
                    nat.nnat.inv = "Status",
                    "CGRC-01, 2022" = 2022, 
                    "CGRC-01, 2023" = 2023, 
                    "CGRC-02, 2022" = 2022, 
                    "CGRC-02, 2023" = 2023, 
                    "CGRC-03, 2022" = 2022,
                    "CGRC-03, 2023" = 2023,
                    "CGRC-05, 2022" = 2022,
                    "CGRC-05, 2023" = 2023) %>% 
  autofit() %>% 
  fit_to_width(max_width = 7)

```

